Topological Equivariant Artist Model for Visualization Library Architecture

### 2022-09-14 paper/main.tex
summary tables

\subsection{Field Type Structure} 
\label{sec:related-work:equivariance}
As with topology, Bertin \cite{bertinSemiologyGraphicsDiagrams2011} codified the expectation that a data value and its visual encoding are expected to have compatible structure. The structure on values has traditionally been described  using the measurement scales-i.e. nominal, ordinal, interval, ratio - which Steven's classified by their mathematical group structure\cite{stevensTheoryScalesMeasurement1946a} and others have formally expanded to include more types \cite{leaFormalizationMeasurementScale, thomasMathematizationNotMeasurement2014}. Generally the structure on each field of the dataset is preserved separately, i.e. each column in a table is mapped to a distinct encoding. Two mathematical concepts for describing a function that preserves structure are \textit{equivariance} and \textit{homomorphism}. An \textit{equivariant} function preserves symmetric group structure on its input and output, while a \textit{homomorphic} function $f$ preserves binary operations on input and output of the same algebraic type. Group structure and binary operations are both describing families of functions or operations that can be applied to data and visual encodings. A group $G$ is a set with an operator $\circ$. The set contains an identity element $e$ and the operator $\circ$ is closed, associative and invertible. Applying elements of a set, such as a group, to another set, such as data $X$, is called an \textit{action}.
\begin{definition}
  An \textcolor{action}{\textbf{action}} $act: G \times X \rightarrow X$ is a function that   satisfies the axioms of identity $act(e, x)$ for $e \in G$ and associativity $act(g, act(f, x)) = act(f \circ g, x)$ for $f,g \in G$ \cite{nlab:action}.
\end{definition}
Given a group $G$ that acts on input $X$ and output $Y$ and a function $f: X \rightarrow Y$, 
 \begin{definition}
$f$ is \textbf{equivariant} when $f(g(x)) = g(f(x))$ for all $g$ in $G$ and for all $x$ in $X$\cite{nlab:equivariant}.
 \begin{definition}
 $f$ is \textbf{homomorphic} when  $f(x_1 \oslash x_2) = f(x_1) \oslash f(x_2)$ for every pair $x_1, x_2$ in $X$\cite{grimaldiDiscreteCombinatorialMathematics2006}. 
 \end{definition}
For example, nominal measurements are permutable, meaning that they are acted on by elements of a permutation group. For the visual transformation to be equivariant, that nominal measurement must be mapped to a visual encoding that is also permutable, such as distinct marker shapes or color hues. We evaluate whether the encoding is homomorphic when the visual encoding does not have a group structure. For example, partial orders are not invertible; therefore a visual encoding of partially ordered data is structure preserving when the ordering is mapped to equivalently ordered visual encodings.

@@

\begin{table}[h!]
  \centering
  {\renewcommand{\arraystretch}{1.2}
  \begin{tabular}{|r c c c l|}
    \hline
 Topological Space  & Data  & Visual  & Graphic  & Encodes \\
   \hline
\textcolor{total}{total space} & \dtotal & \vtotal & \gtotal & fields + topology \\
\textcolor{base}{base space} & \dbasec & \dbasec & \gbasec & topological structure\\
\textcolor{base}{open set} & \opensetc & \opensetc & \opensetgc & subspace \\
\textcolor{fiber}{fiber space} & \dfiberc & \vfiberc & \gfiberc  & field structure\\
\hline
  \end{tabular}
  \caption{Topological spaces introduced in \autoref{sec:atct:fiber-bundles} and \autoref{sec:construction:vtotal}}
  \label{}
\end{table}

\begin{table}[h!]
  \centering
  {\renewcommand{\arraystretch}{1.2}
  \begin{tabular}{|r r r r r|}
     & topological structure & subspace & field structure &  fields + topology\\
     & \textcolor{base}{base space} & \textcolor{base}{openset} & \textcolor{fiber}{field types} & \textcolor{total}{total space}
  \end{tabular}
  \caption{Topological spaces introduced in \autoref{sec:atct:fiber-bundles} and \autoref{sec:construction:vtotal}}
  \label{tab:appendix:summary:objects}
\end{table}

\begin{table}[h!]
  \centering
  {\renewcommand{\arraystretch}{1.2}
  \begin{tabular}{|r c c c l|}

\textcolor{section}{section} & $\dsectionc: \dbase \textcolor{section}{\rightarrow} \dfiberc$ &  $\vsectionc: \dbase \textcolor{section}{\rightarrow} \vfiberc$ & $\gsectionc: \gbase \textcolor{section}{\rightarrow} \gfiberc$ & data set \\
\textcolor{sheaf}{sheaf} & $\sheafc_{\dbasec, \dtotalc}: \opensetc \rightarrow \cgamma{\opensetc}{\dtotalc\restriction_{\opensetc}}$ & $\sheafc_{\dbasec, \vtotalc}: \opensetc \rightarrow \cgamma{\opensetc}{\vtotalc\restriction_{\opensetc}}$ & $\sheafc_{\gbasec, \gtotalc}: \opensetgc \rightarrow \cgamma{\opensetc}{\gtotalc\restriction_{\opensetgc}}$ & subspace to records map \\ 
\hline
  \end{tabular}
  \caption{Functions that associate topological subspaces with records. Equivalent constructions can be applied to elements in visual and graphic sheaves, and these functions are distinguised through subscripts $\dfunc_{\dtotal}$, $\dfunc_{\vtotal}$ and $\dfunc_{\gtotal}$}
  \label{tab:appendix:summary:datafunctions}
\end{table}

\begin{table}[h!]
\centering
{\renewcommand{\arraystretch}{1.2}
\caption{Functions $\dfunc=(\dfunchc, \dfunctc)$ for modifying data records}
\label{tab:appendix:summary:datamod}
\end{table}

\note{color and replace source/target with constraint}
\begin{table}[h!]
  \centering
\begin{tabular}{|r|l|l|}
  \hline
      & function & constraint \\
  \hline
  \textcolor{artist}{artist} & $\vartist: \Gamma(\dbase, \dtotal) \rightarrow \Gamma(\gbase, \gtotal)$ &  \\
  \hline 
  \textcolor{functor}{lookup} & $\vindex: \gbase\rightarrow \dbase$  &\\
  \multicolumn{4}{c}{}\\
  \hline
  \textcolor{artist}{encoder} & $\vchannel: \Gamma(\dbase, \dtotal) \rightarrow \Gamma(\dbase, \vtotal) $  & \\
  \hline
  \textcolor{artist}{compositor} & $\vmark: \Gamma(\dbase,\vtotal) \rightarrow \Gamma(\gbase, \vtotal)$ &  \\
  \hline
\end{tabular}
\caption{artist, verification functions, and construction $\vartist = \vmarkc \circ \nu$ introduced in \autoref{sec:atct:xi}, \autoref{sec:artist}, and \autoref{sec:construction}}
\label{tab:appedendix:summary:artist}
\end{table}

>>>>

\subsection{Field Type Structure} 
\label{sec:related-work:equivariance}
As with topology, Bertin \cite{bertinSemiologyGraphicsDiagrams2011} codified the expectation that a data value and its visual encoding are expected to have compatible structure. The structure on values has traditionally been described  using the measurement scales-i.e. nominal, ordinal, interval, ratio - which Steven's classified by their mathematical group structure\cite{stevensTheoryScalesMeasurement1946a} and others have formally expanded to include more types \cite{leaFormalizationMeasurementScale, thomasMathematizationNotMeasurement2014}. Generally the structure on each field of the dataset is preserved separately, i.e. each column in a table is mapped to a distinct encoding. Two mathematical concepts for describing a function that preserves structure are \textit{equivariance} and \textit{homomorphism}. An \textit{equivariant} function preserves symmetric group structure on its input and output, while a \textit{homomorphic} function $f$ preserves binary operations on input and output of the same algebraic type. Group structure and binary operations are both describing families of functions or operations that can be applied to data and visual encodings. A group $G$ is a set with an operator $\circ$. The set contains an identity element $e$ and the operator $\circ$ is closed, associative and invertible. Applying elements of a set, such as a group, to another set, such as data $X$, is called an \textit{action}.
\begin{definition}
  An \textcolor{action}{\textbf{action}} $act: G \times X \rightarrow X$ is a function that   satisfies the axioms of identity $act(e, x)$ for $e \in G$ and associativity $act(g, act(f, x)) = act(f \circ g, x)$ for $f,g \in G$ \cite{nlab:action}.
\end{definition}
Given a group $G$ that acts on input $X$ and output $Y$ and a function $f: X \rightarrow Y$, 
 \begin{definition}
$f$ is \textbf{equivariant} when $f(g(x)) = g(f(x))$ for all $g$ in $G$ and for all $x$ in $X$\cite{nlab:equivariant}.
 \end{definition}
 \begin{definition}
 $f$ is \textbf{homomorphic} when  $f(x_1 \oslash x_2) = f(x_1) \oslash f(x_2)$ for every pair $x_1, x_2$ in $X$\cite{grimaldiDiscreteCombinatorialMathematics2006}. 
 \end{definition}
For example, nominal measurements are permutable, meaning that they are acted on by elements of a permutation group. For the visual transformation to be equivariant, that nominal measurement must be mapped to a visual encoding that is also permutable, such as distinct marker shapes or color hues. We evaluate whether the encoding is homomorphic when the visual encoding does not have a group structure. For example, partial orders are not invertible; therefore a visual encoding of partially ordered data is structure preserving when the ordering is mapped to equivalently ordered visual encodings.

@@

\begin{table}[h!]
  \centering
  {\renewcommand{\arraystretch}{1.5}
  \begin{tabular}{|r | c c c|}
    \hline
    &\textcolor{base}{point}/\textcolor{base}{openset}/\textcolor{base}{base space} & \textcolor{fiber}{fiber space} & \textcolor{total}{total space}\\
     &  location/subset/topology & fields &  dataset type\\
    \hline
   Data & $\dbasepoint \in \opensetc \subseteq \dbasec$ & \dfiberc & \dtotalc\\
   Visual & $\dbasepoint \in \opensetc \subseteq \dbasec$  & \vfiberc & \vtotalc\\
   Graphic & $\gbasepoint \in \opensetgc \subseteq \gbasec$ & \gfiberc & \gtotalc\\
   \hline
  \end{tabular}
  \caption{Topological spaces introduced in \autoref{sec:atct:fiber-bundles}}
  \label{tab:appendix:summary:objects}
  }
\end{table}

\begin{table}[h!]
  \centering
  {\renewcommand{\arraystretch}{1.5}
  \begin{tabular}{|r | l l | }
    \hline
     & \textcolor{section}{section} & \textcolor{sheaf}{sheaf} \\
     & location to record &  subset to set of possible records  \\
     \hline
  Data & $ \cgamma{\dbasec}{\dtotalc} \ni \dsectionc: \dbase \textcolor{section}{\rightarrow} \dfiberc$ & $\sheafc_{\dbasec, \dtotalc}: \opensetc \rightarrow \cgamma{\opensetc}{\dtotalc\restriction_{\opensetc}}$\\
  Visual &  $\cgamma{\dbasec}{\vtotalc} \ni \vsectionc: \dbase \textcolor{section}{\rightarrow} \vfiberc$ & $\sheafc_{\dbasec, \vtotalc}: \opensetc \rightarrow \cgamma{\opensetc}{\vtotalc\restriction_{\opensetc}}$\\
  Graphic &    $\cgamma{\gbasec}{\gtotalc} \ni \gsectionc: \gbase \textcolor{section}{\rightarrow} \gfiberc$ &  $\sheafc_{\gbasec, \gtotalc}: \opensetgc \rightarrow \cgamma{\opensetc}{\gtotalc\restriction_{\opensetgc}}$ \\ 
  \hline
  \end{tabular}
  \caption{Functions that associate topological subspaces with records, discussed in \autoref{sec:atct:fb:sections} and \autoref{sec:atct:sheaves}}
  \label{tab:appendix:summary:datafunctions}
  }
\end{table}

\begin{table}[h!]
  \centering
  {\renewcommand{\arraystretch}{1.5}
  \begin{tabular}{|r | l  l |}
\hline
 \gbasepointc\ to \dbasepointc& $ \vindexc: \opensetgc \rightarrow \opensetc$ & for $\dbasepointc \in \opensetc$ exists $\gbasepointc \in \opensetgc$ \\
 & & s.t. $\vindexc(\gbasepointc) = \dbasepointc$\\
 graphic for \dbasepointc & $ \vindexpushc \gsectionc: \opensetc \rightarrow \vindexpushc\gtotalc\restriction_{\opensetc}$ & $\vindexpushc\gsectionc(\dbasepointc)(\gbasepointc) = \gsectionc(\gbasepointc)$ \\
data for \gbasepointc & $\vindexpullc\dsectionc: \opensetgc \rightarrow  \vindexpullc\dtotalc\restriction_{\opensetgc}$ & $\vindexpullc\dsectionc(\gbasepointc)=\dsectionc(\vindexc(\gbasepointc)) = \dsectionc(\dbasepointc)  \\
\hline
  \end{tabular}
  \caption{Functors between graphic and data base spaces  \autoref{sec:atct:xi}}
  \label{tab:appendix:summary:transport}
  }
\end{table}

\begin{table}[h!]
\centering
{\renewcommand{\arraystretch}{1.2}
\begin{tabular}{}
  \note{changing locations, moving records to new locations, changing records}

\end{tabular}

\caption{Functions $\dfunc=(\dfunchc, \dfunctc)$ for modifying data records. Equivalent constructions can be applied to elements in visual and graphic sheaves, and these functions are distinguised through subscripts $\dfunc_{\dtotal}$, $\dfunc_{\vtotal}$ and $\dfunc_{\gtotal}$}
\label{tab:appendix:summary:datamod}
}
\end{table}

\note{color}
\begin{table}[h!]
  \centering
  {\renewcommand{\arraystretch}{1.2}
\begin{tabular}{|r|l|l|}
  \hline
      & function & constraint \\
  \hline
  \textcolor{artist}{artist} & $\vartist: \Gamma(\dbase, \dtotal) \rightarrow \Gamma(\gbase, \gtotal)$ &  \\
  \hline 
  \textcolor{functor}{lookup} & $\vindex: \gbase\rightarrow \dbase$  &\\
  \hline
  \textcolor{artist}{encoder} & $\vchannel: \Gamma(\dbase, \dtotal) \rightarrow \Gamma(\dbase, \vtotal) $  & \\
  \hline
  \textcolor{artist}{compositor} & $\vmark: \Gamma(\dbase,\vtotal) \rightarrow \Gamma(\gbase, \vtotal)$ &  \\
  \hline
\end{tabular}
\caption{artist, verification functions, and construction $\vartist = \vmarkc \circ \nu$ introduced in \autoref{sec:artist}, and \autoref{sec:construction}}
\label{tab:appedendix:summary:artist}
}
\end{table}
